A Logical Solution of the Linearity Challenge

### 2025-12-03 paper/DeadlockFreedom.lagda.tex
revised DF

As we have seen in \Cref{sec:calculus}, the deadlock freedom result
(\Cref{thm:df}) rests on some notions and predicates about processes which we
now formalize in Agda. First of all we need to distinguish \emph{threads} from
\emph{cuts}. As it turns out, in the formalization it is convenient to provide
an even more fine-grained classification of threads, distinguishing between
links, input/output actions on linear/shared channels, and sometimes also
whether such actions operate on free or bound channels. We define predicates for
each of these process classes:

@@

These predicates are relatively uninteresting since they are essentially
isomorphic to the relevant parts of the \AgdaDatatype{Process} datatype. The
only remarkable aspect is that \AgdaDatatype{Input}, \AgdaDatatype{Output},
\AgdaDatatype{Client} and \AgdaDatatype{Server} the channel being acted upon by
the thread is the \emph{first} in $\Context$, hence it is the \emph{most
recently introduced} channel, whereas in \AgdaDatatype{Delayed} and
\AgdaDatatype{DelayedServer} the channel is older. This allows us to distinguish
those threads that, in the context of a cut, operate on the channel bound by the
cut or on a free channel. As an illustration, consider the
\AgdaInductiveConstructor{wait} case in \AgdaDatatype{Input} and in
\AgdaDatatype{Delayed}. In the former predicate we have

@@

\begin{code}
  wait : ∀{Γ Δ P} (p : Γ ≃ [] + Δ) → Input (wait (< p) P)
\end{code}
where the \AgdaInductiveConstructor{split-l} constructor indicates that the
thread operates on the most recent channel. In the latter predicate we have

@@

\begin{code}
  wait     : ∀{C Γ Δ P} (p : Γ ∋ ⊥ ⊳ Δ) → Delayed (wait (>_ {C} p) P)
\end{code}
where the \AgdaInductiveConstructor{split-r} constructor indicates that the
thread operates on a channel other than the most recent one.

@@

In order to prove that every (well-typed) process is alive, it is convenient to
also define a ``canonical'' form for cuts, which is a form that matches at least
one of the left-hand-sides of the rules for structural pre-congruence or
reduction (\Cref{tab:semantics}). This is the notion where the fine-grained
classification of threads introduced earlier comes into play.

@@

A \emph{canonical cut} $\Cut\x{P}{Q}$ has one of these forms:
\begin{itemize}
\item $P$ is a link (\AgdaInductiveConstructor{cc-link}), or
\item $P$ operates on a channel other than $x$ and is not a server
(\AgdaInductiveConstructor{cc-delayed}), or
\item $P$ and $Q$ are servers and $P$ operates on a channel other than $x$
(\AgdaInductiveConstructor{cc-delayed-server}), or
\item $P$ performs an output on $x$ and $Q$ performs an input on $x$
(\AgdaInductiveConstructor{cc-redex} and \AgdaInductiveConstructor{cc-connect}).
\end{itemize}

@@

Every cut $\Cut\x{P}{Q}$ where both $P$ and $Q$ are threads can be rewritten
into a canonical cut using structural pre-congruence:

@@

When $P$ is a thread, then it is obviously observable and hence alive. When $P$
is a cut $\Cut\x{P}{Q}$, \AgdaFunction{deadlock-freedom} is applied recursively
on $P$ and $Q$. If either of these applications yields a reduction, then the
whole cut is reducible and therefore alive. If both applications yield a thread,
then we conclude that the cut is alive first rewriting it into a canonical cut
(\AgdaFunction{canonical-cut}) and then using
\AgdaFunction{canonical-cut-alive}.

>>>>

As we have seen in \Cref{sec:calculus}, the deadlock freedom result
(\Cref{thm:df}) rests on some notions and predicates about processes which we
now formalize in Agda. First of all we need to distinguish \emph{threads} from
\emph{cuts}. As it turns out, in the formalization it is convenient to provide a
more fine-grained classification of threads, distinguishing between links,
input/output actions on linear/shared channels, and sometimes also whether such
actions operate on free or bound channels. We define predicates for each of
these process classes:

@@

These predicates are relatively uninteresting since they are essentially
isomorphic to the relevant parts of the \AgdaDatatype{Process} datatype. The
only remarkable aspect is that in \AgdaDatatype{Input}, \AgdaDatatype{Output},
\AgdaDatatype{Client} and \AgdaDatatype{Server} the channel being acted upon by
the thread is the \emph{first} in $\Context$, hence it is the \emph{most
recently introduced} channel, whereas in \AgdaDatatype{Delayed} and
\AgdaDatatype{DelayedServer} the channel is older. This allows us to distinguish
those threads that, in the context of a cut, operate on the channel bound by the
cut or on a free channel. As an illustration, consider the
\AgdaInductiveConstructor{wait} case in \AgdaDatatype{Input} and in
\AgdaDatatype{Delayed}. In the former predicate we have

@@

\begin{code}
  wait : ∀{Γ Δ P} (p : Γ ≃ [] + Δ) → Input (wait (< p) P)
\end{code}
where the \AgdaInductiveConstructor{<} constructor indicates that the thread
operates on the most recent channel. In the latter predicate we have

@@

\begin{code}
  wait     : ∀{C Γ Δ P} (p : Γ ∋ ⊥ ⊳ Δ) → Delayed (wait (>_ {C} p) P)
\end{code}
where the \AgdaInductiveConstructor{>} constructor indicates that the thread
operates on a channel other than the most recent one. Note that here we have to
specify the type $C$ in front of $\Context$ or else Agda is unable to resolve
some metavariables.

@@

In order to prove that every (well-typed) process is alive, it is convenient to
also define a ``canonical'' form for cuts, that is a form that matches at least
one of the left-hand-sides of the rules for structural pre-congruence or
reduction (\Cref{tab:semantics}). This is the notion where the fine-grained
classification of threads introduced earlier comes into play.

@@

A \emph{canonical cut} $\Cut[A]\x{P}{Q}$ has one of these forms:
\begin{itemize}
\item $P$ is a link (\AgdaInductiveConstructor{cc-link}), or
\item $P$ operates on a channel other than $x$ and is not a server
(\AgdaInductiveConstructor{cc-delayed}), or
\item $P$ and $Q$ are servers and $P$ operates on a channel other than $x$
(\AgdaInductiveConstructor{cc-delayed-server}), or
\item $P$ performs an output on $x$ and $Q$ performs an input on $x$
(\AgdaInductiveConstructor{cc-redex} and \AgdaInductiveConstructor{cc-connect}).
\end{itemize}

@@

Every cut $\Cut[A]\x{P}{Q}$ where both $P$ and $Q$ are threads can be rewritten
into a canonical cut using structural pre-congruence:

@@

When $P$ is a thread, then it is obviously observable and hence alive. When $P$
is a cut $\Cut[A]\x{P}{Q}$, \AgdaFunction{deadlock-freedom} is applied
recursively on $P$ and $Q$. If either of these applications yields a reduction,
then the whole cut is reducible and therefore alive. If both applications yield
a thread, then we conclude that the cut is alive first rewriting it into a
canonical cut (\AgdaFunction{canonical-cut}) and then using
\AgdaFunction{canonical-cut-alive}.
